### 2019-07-10 bp_discussion.tex
addressed concerns of Reviewer 3
modified intro and discussion to frame it more towards HCI

%!TEX root = bioPrediction_main.tex
\section{Discussion}

\vspace{0.05in}
\noindent
\textbf{Target Workspaces:}
Currently, the cost of biometric sensors and necessary infrastructure, such as automated light and sound systems for adjusting the environment, makes our approach most appropriate for high-value workspaces, such as control rooms, command centers, or dispatch offices. However, as standard office settings become more personalizable (e.g., via adjustable desks, lighting, and sound showers) and sensor costs decrease, our approach could be applied to any office environment, and thus could impact a large percentage of modern workers. As in modern cars, temperature and lighting could be regulated on a per-person basis, which would allow the environment to react to the person's current state and to maximize each person's preferences and productivity (e.g., preferences of men and women in temperature~\cite{Karjalainen07}). Further research would be needed to identify how to balance needs across a group of office workers and how to handle conflicting levels between different group members.

\vspace{0.05in}
\noindent
\textbf{Imbalanced Data:}
Study participants provided highly imbalanced data in their survey responses, with most participants only taking advantage of a subset of the Likert-scale values and the data points mostly being clustered around the middle of the scale, as can be seen in Table~\ref{responseDistribution}. While some of the imbalance is expected due to certain classes, such as 'not stressed', being more common in the workplace, this imbalance also provides challenges in the training and assessment of a machine learning classifier, as also found by others, e.g.~\cite{Exler16}. We addressed this for the training by oversampling in case of few data samples for the individual models and undersampling in case of a general model where more data was available. Especially in light of this imbalance in the data, the results we achieved with our models are encouraging. For the assessment of the classifiers' performance we addressed the imbalance by not just presenting accuracy, but also by focusing on prediction and recall to examine the classifier's performance in predicting the infrequent (yet more important) cases, such as when a user is struggling to stay awake and an intervention or warning might be needed most. 
% \textbf{Unbalanced Data:}
% As shown in Table~\ref{responseDistribution}, the typical participant provided highly unbalanced survey responses, with most responses exhibiting a central tendency. Many participants provided zero or one very high (or very low) value on the Likert scale. In spite of this, our results show that by using biometric sensors and leveraging a machine learning approach, we are able to predict stress, focus, and awakeness better than the baseline, with improvements up to 84.64\%. Taking into account that the model correctly predicts many cases that happen infrequently, this is especially promising. These infrequent cases, such as when a user is highly stressed or struggling to stay awake, are in fact the most important cases to detect, as they are when intervention is needed most.

\vspace{0.05in}
\noindent
\textbf{Ground Truth and Self-Reporting:} 
One of the key points for developing a good classifier for awakeness, focus and stress, is to collect a valid ground truth to be used as the output measure. When designing the study, we therefore spent an extensive amount of time on determining the exact questions to ask in the experience sampling, consulting experts in the area, and basing the questions and wording on previous research and studies. Yet, the reliability and validity of self-reports have been questioned in the past due to subjective biases, lack of care in reporting, and the highly individual nature of reporting aspects such as stress~\cite{Hernandez11,Hovsepian15}. In addition, in contexts such as the workplace, employees might be afraid to genuinely report levels of aspects, such as sleepiness. Hence, there is a chance that the collected experience samples do not adequately reflect the ground truth of the underlying variable under investigation. It could even be the case that certain biometrics might represent a more accurate ground truth of the studied phenomena than the self-reports. This suggests that a more confirmative study rather than an inquiry study could be a better approach, and we will explore such routes in future work.

\vspace{0.05in}
\noindent
\textbf{Corpus Size:}
In this study we collected data from 14 professionals over the course of eight weeks. By scholarly standards this constitutes a large dataset --- especially given that the participants were not students. However, in the context of machine learning this corpus is relatively small. Thus, the model's overall performance, while promising, is only an indication of the performance that a larger corpus could provide. Our results show that there is a positive correlation between data added to the training corpus and the performance level of the model. This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis.

@@

\vspace{0.05in}
\noindent
\textbf{Per-User Model vs. Across-Users Model:}
Our results show that individually trained models outperform on average a more general model. Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), and thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end.

>>>>

%!TEX root = bioPrediction_main.tex
\section{Discussion}
%\vspace{0.05in}
\noindent
\textbf{Implications for HCI Design and Workspaces:}
The results discussed in this paper establish great potential
to increase the well-being of knowledge workers and the quality of their work
through monitoring their biometric signals and computer interactions with the
goal of recognizing their stress, awakeness, and focus levels in an uninvasive 
and automatic manner. Being able to accurately recognize periods of high-
stress in workers enables companies to prevent or de-escalate potentially 
dangerous situations in the workplace, e.g. confrontation between co-workers; 
or it can help to design more team-based approaches such as an 
awareness dashboard of a teams’ stress level.
Similarly, awakeness and focus
can be enhanced given such a technology
e.g. through individual approaches such as adapting the lighting in a control 
room, or scheduling breaks to prevent focus loss.

Currently, the cost of biometric sensors and necessary infrastructure, such as automated light and sound systems for adjusting the environment, makes our approach most appropriate for high-value workspaces, such as control rooms, command centers, or dispatch offices. However, as standard office settings become more personalizable (e.g., via adjustable desks, lighting, and sound showers) and sensor costs decrease, our approach could be applied to any office environment, and thus could impact a large percentage of modern workers. As in modern cars, temperature and lighting could be regulated on a per-person basis, which would allow the environment to react to the person's current state and to maximize each person's preferences and productivity (e.g., preferences of men and women in temperature~\cite{Karjalainen07}). Further research would be needed to identify how to balance needs across a group of office workers and how to handle conflicting levels between different group members.

%\vspace{0.05in}
\noindent
\textbf{Imbalanced Data:}
Study participants provided highly imbalanced data in their survey responses, with most participants only taking advantage of a subset of the Likert-scale values and the data points mostly being clustered around the middle of the scale, as can be seen in Table~\ref{responseDistribution}. While some of the imbalance is expected due to certain classes, such as 'not stressed', being more common in the workplace, this imbalance also provides challenges in the training and assessment of a machine learning classifier, as also found by others, e.g.~\cite{Exler16}. We addressed this for the training by oversampling in case of few data samples for the individual models and undersampling in case of a general model where more data was available. Especially in light of this imbalance in the data, the results we achieved with our models are encouraging. For the assessment of the classifiers' performance we addressed the imbalance by not just presenting accuracy, but also by focusing on prediction and recall to examine the classifier's performance in predicting the infrequent (yet more important) cases, such as when a user is struggling to stay awake and an intervention or warning might be needed most. 
% \textbf{Unbalanced Data:}
% As shown in Table~\ref{responseDistribution}, the typical participant provided highly unbalanced survey responses, with most responses exhibiting a central tendency. Many participants provided zero or one very high (or very low) value on the Likert scale. In spite of this, our results show that by using biometric sensors and leveraging a machine learning approach, we are able to predict stress, focus, and awakeness better than the baseline, with improvements up to 84.64\%. Taking into account that the model correctly predicts many cases that happen infrequently, this is especially promising. These infrequent cases, such as when a user is highly stressed or struggling to stay awake, are in fact the most important cases to detect, as they are when intervention is needed most.

%\vspace{0.05in}
\noindent
\textbf{Ground Truth and Self-Reporting:} 
One of the key points for developing a good classifier for awakeness, focus and stress, is to collect a valid ground truth to be used as the output measure. When designing the study, we therefore spent an extensive amount of time on determining the exact questions to ask in the experience sampling, consulting experts in the area, and basing the questions and wording on previous research and studies. Yet, the reliability and validity of self-reports have been questioned in the past due to subjective biases, lack of care in reporting, and the highly individual nature of reporting aspects such as stress~\cite{Hernandez11,Hovsepian15}. In addition, in contexts such as the workplace, employees might be afraid to genuinely report levels of aspects, such as sleepiness. Hence, there is a chance that the collected experience samples do not adequately reflect the ground truth of the underlying variable under investigation. It could even be the case that certain biometrics might represent a more accurate ground truth of the studied phenomena than the self-reports. This suggests that a more confirmative study rather than an inquiry study could be a better approach, and we will explore such routes in future work.

%\vspace{0.05in}
\noindent
\textbf{Corpus Size:}
In this study we collected data from 14 professionals over the course of eight weeks. By scholarly standards this constitutes a large dataset --- especially given that the participants were not students. However, in the context of machine learning this corpus is relatively small. Thus, the model's overall performance, while promising, is only an indication of the performance that a larger corpus could provide. Our results show that there is a positive correlation between data added to the training corpus and the performance level of the model. This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis.

@@

%\vspace{0.05in}
\noindent
\textbf{Per-User Model vs. Across-Users Model:}
Our results show that individually trained models outperform on average a more general model. Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), and thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end.

### 2019-07-10 bp_intro_v3.tex
modified intro and discussion to frame it more towards HCI

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether business or non-business) will be its knowledge workers and their productivity'}~\cite{drucker1999knowledge}. Yet, in today's workplaces, knowledge workers and their time are not necessarily treated as the most valuable asset, and they constantly face challenges, such as a high work fragmentation, continuous disruptions and distractions, highly complex and demanding tasks, and long working hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. These challenges make it difficult for a knowledge worker to stay focused and can impact both the quality of work performed and productivity. For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and a lower performance~\cite{bailey2001effects,mark2008cost}. In some cases,
the consequences are even more severe; in control room
settings, the consequences can be accidents that cost millions of dollars in losses. 
%when a sleepy operator might, for example, miss early warning signs and cause...
%\todo{Nick: can you find a good example here and citation?}. 
%Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}. 
Less immediate, but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems. It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

Being able to continuously measure human aspects --- such as focus, awakeness, and stress --- in the workplace bears huge potential for better supporting knowledge workers, combatting workplace challenges, and improving knowledge worker productivity. Such measures could be used to provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to automatically adjust lighting to reduce sleepiness among control room operators, or to continuously monitor stress levels to trigger pro-active interventions and reduce health problems.

Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure such human aspects. A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive state 
%This change is regarding Reviewer 2's concern: what are the "aspects"?
(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002} 
and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Yet, very few 
studies measure productivity-related factors in the workplace over an 
extended period of time, and little is known about measuring the more 
abstract concept of focus (directed attention). 
% Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
Given the importance of focus, described as a 
combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also 
deserves further study.

The objective of our research is to develop continuous (and automatic) measures of focus, awakeness, and stress in the workplace. These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future. For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions during an extended period of high stress.

Our work builds upon previous work and extends it by performing an eight-week study in the workplace with 14 knowledge workers, collecting biometric and computer interaction data as well as experience samples on focus, awakeness, and stress. Our participants, who perform various job functions for a research and development group within a single large corporation, wore a single biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} with low invasiveness that captured heart- and skin-related measures. This modality was chosen to ease longitudinal deployment in the field. Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects. In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed how the combination
of these two training sets compares to the performance of each
set individually.

The results of our analysis show that biometrics of a single minimally invasive sensor can be used to predict productivity-related indicators accurately, with the abstract concept of focus (predictably) being the hardest to detect. The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals. Our results further show that overall combining biometric features with computer interaction information 
results in an improvement
over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to possible target workspaces, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

>>>>

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether 
business or non-business) will be its knowledge workers and their 
productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
challenges, such as a high work fragmentation, continuous disruptions and 
distractions, highly complex and demanding tasks, and long working 
hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. These 
challenges make it difficult for a knowledge worker to stay focused and
impact interaction with their digital environment, the quality of work performed and productivity. For instance, the 
continuous disruptions that knowledge workers face in office work environments 
can lead to a higher error rate and a lower 
performance~\cite{bailey2001effects,mark2008cost}. In some cases,
the consequences are even more severe; e.g., in control room
settings, the consequences of an erroneous intraction can lead to accidents valued in millions of dollars in losses. 
%when a sleepy operator might, for example, miss early warning signs and cause...
%\todo{Nick: can you find a good example here and citation?}. 
%Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}. 
Less immediate, but just as insidious, stress at the workplace is a growing 
concern and one of the most common work-related health problems. It leads to 
fatigue, burnout and various other illnesses, ultimately resulting in work 
absences and marked productivity 
losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

Being able to continuously measure human aspects --- such as focus, awakeness, 
and stress --- in the workplace bears huge potential for better supporting 
knowledge workers combatting workplace challenges, and improving productivity 
and well-being. Such measures could be used to provide better interaction with 
their environment, such as management 
of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to 
automatically adjust lighting and temperature to reduce sleepiness, or to continuously monitor stress levels to trigger pro-active 
interventions and reduce health problems.

Fortunately, with recent advances in sensing technologies there is increasing 
hope that we can accurately and automatically measure such human aspects. A 
number of HCI studies have discussed and investigated the use of biometrics to 
measure aspects such as a person's cognitive state 
%This change is regarding Reviewer 2's concern: what are the "aspects"?
(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002}, 
and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Similarly,
studies have examined the occurrence, cause and effect of similar 
aspects in various contexts~\cite{mark2014bored,mark2008cost}, how these can be measured~\cite{Hernandez14}, and 
 how they might be modulated/regulated~\cite{Snyder15}. 
While previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether and how these three variables ---stress, focus and awakeness--- can be measured over an extended period of time in a knowledge worker's context, 
%Yet, very few 
%studies measure productivity-related factors in the workplace over an 
%extended period of time
and little is known about measuring the more 
abstract concept of focus (i.e., directed attention). 
% Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
Given the importance of focus, described as a 
combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also 
deserves further study.

The objective of our research is to develop continuous (and automatic) 
measures of focus, awakeness, and stress in the workplace. These measures are 
targeted towards improving knowledge workers' productivity and 
well-being in the future by customizing interaction with their environment; e.g., by automatically protecting the workers
from audible interruptions during a high focus period by showering them 
with white noise or by recommending stress-reducing interventions during an 
extended period of high stress.

Our work builds upon previous work and extends it by performing an eight-week 
study in the workplace with 14 knowledge workers, collecting biometric and 
computer interaction data as well as experience samples on focus, awakeness, 
and stress. Our participants, who perform various job functions for a research 
and development group within a single large corporation, wore a single 
biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} 
with low invasiveness that captured heart-, respiration- and skin-related measures. This 
modality was chosen to ease longitudinal deployment in the field. Using 
machine learning, we created classifiers and analyzed their ability to predict 
the three productivity-related human aspects. Similarly, we captured computer 
interaction features from participants' everyday computer usage and
analyzed how a model trained using the combination
of these two training sets compares to the performance of models using each
set individually.

The results of our analysis show that biometrics of a single minimally invasive sensor can be used to predict productivity-related indicators accurately, with the abstract concept of focus (predictably) being the hardest to detect. The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals. Our results further show that overall combining biometric features with computer interaction information 
results in an improvement
over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}
